---
title: "Poisson Regression"
subtitle: "Zero-inflated Poisson models"
author: "Prof. Maria Tackett"
date: "02.07.22"
output:
  xaringan::moon_reader:
    #mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta310-slides.css"
    logo: sta310-sticker.png
    lib_dir: libs/font-awesome
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%" 
      ratio: "16:9"
    self_contained: true
---

```{r setup, include = F}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(GGally)
library(xaringanExtra)
library(knitr)
library(patchwork)
library(viridis)
library(ggfortify)
library(gridExtra)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: middle, center

## [Click for PDF of slides](09-poisson-pt4.pdf)

---


## Announcements

- For Mon, Feb 14: BMLR - [Chapter 5: Generalized Linear Models: A Unifying Theory](https://bookdown.org/roback/bookdown-BeyondMLR/ch-glms.html)


- [Mini-Project 01](https://sta310-sp22.github.io/assignments/projects/mini-project-01.html): 
  - Presentation due  **Wed, Feb 09 at 3:30pm**
  - Final write up & GitHub repo organization due **Fri, Feb 11 at 5pm**

- [HW 02](https://sta310-sp22.github.io/assignments/hw/hw-02.html) **due TODAY at 11:59pm**

---

class: middle

## HW 02 questions? 

---

class: middle

## Presentation order

(all presentations and write ups due on Wed, Feb 09 at 3:30pm)


---

## Learning goals 

- Fit and interpret the Zero-inflated Poisson model 

- Write likelihood for Poisson and Zero-inflated Poisson model

---

## Data: Weekend drinking 

The data [`weekend-drinks.csv`](data/weekend-drinks.csv) contains information from a survey of 77 students in a introductory statistics course on a dry campus. 

**Variables** 

- `drinks`: Number of drinks they had in the past weekend 
- `off_campus`: 1 - lives off campus, 0 otherwise
- `first_year`: 1 - student is a first-year, 0 otherwise
- `sex`: f - student identifies as female, m - student identifies as male 

**Goal**: The goal is explore factors related to drinking behavior on a dry campus.

.footnote[Case study in BMLR - Section 4.10]

---

## EDA: Response variable

```{r, echo = F}
drinks <- read_csv("data/weekend-drinks.csv")
```


```{r echo = F, out.width = "70%"}
ggplot(data = drinks, aes(x = drinks)) + 
  geom_histogram(aes(x = drinks), binwidth = 1, alpha = 0.8, fill = "gray", color = "gray") + 
  labs(title = "Observed number of drinks", 
       subtitle = "Mean = 2.013")
```

```{r echo = F, eval = F}
drinks %>%
  summarise(mean = mean(drinks), var = var(drinks)) %>%
  kable(digits = 3)
```

---

## Observed vs. expected response

```{r echo = F,  out.width = "70%"}
drinks %>%
  mutate(sim = rpois(77, 2.013)) %>%
ggplot() +   
  geom_histogram(aes(x = drinks), binwidth = 1, alpha = 0.7, fill = "gray", color = "gray") + 
  geom_histogram(aes(x = sim), binwidth = 1, fill = "steelblue", alpha = 0.73, color = "steelblue") + 
  labs(title = "Observed (gray) vs. Expected (blue) in Poisson(2.013)")
```

--

.center[**What does it mean to be a "zero" in this data?**]

---

## Two types of zeros

There are two types of zeros

- Those who happen to have a zero in the data set (people who drink but happened to not drink last weekend)
- Those who will always report a value of zero (non-drinkers)
   - These are called **true zeros** 

--

We introduce a new parameter $\alpha$ for the proportion of true zeros,  then fit a model that has two components:

--


`r emo::ji("one")`  The association between mean number of drinks and various characteristics among those who drink 

`r emo::ji("two")`  The estimated proportion of non-drinkers

---


## Zero-inflated Poisson model

**Zero-inflated Poisson (ZIP)** model has two parts

`r emo::ji("one")` Association, among those who drink, between the mean number of drinks and predictors sex and off campus residence

$$\log(\lambda) = \beta_0 + \beta_1 ~ off\_campus + \beta_2 ~ sex$$
where $\lambda$ is the mean number of drinks among those who drink

--

`r emo::ji("two")` Probability that a student does not drink

$$\text{logit}(\alpha) = \log\Big(\frac{\alpha}{1- \alpha}\Big) = \beta_0 + \beta_1 ~ first\_year$$

where $\alpha$ is the proportion of non-drinkers

--

**Note:** The same variables can be used in each component

---

## Details of the ZIP model

- The ZIP model is a special case of a **latent variable model**
  - A type of **mixture model** where observations for one or more groups occur together but the group membership unknown

- Zero-inflated models are a common type of mixture model; they apply beyond Poisson regression

---

## ZIP model in R

Fit ZIP models using the `zeroinfl` function from the **pscl** R package.

```{r}
library(pscl)

drinks_zip <- zeroinfl(drinks ~ off_campus + sex | first_year,
                data = drinks)
drinks_zip
```

---

## Tidy output

Use the `tidy` function from the  **poissonreg** package for tidy model output.

```{r}
library(poissonreg)
```


**Mean number of drinks among those who drink**

```{r}
tidy(drinks_zip, type = "count") %>% kable(digits = 3)
```

---

## Tidy output

**Proportion of non-drinkers**

```{r}
tidy(drinks_zip, type = "zero") %>% kable(digits = 3)
```



---

## Interpreting the model coefficients

```{r echo = F}
tidy(drinks_zip, type = "count") %>% kable(digits = 3)
```

<br> 

.question[
- Interpret the intercept.
- Interpret the coefficients of `off_campus` and `sexm`.

[Click here](https://forms.gle/vkV71Jp1DuocfgPBA) to submit our response. 
]


```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0,
          margin = "5%")
```


---

## Estimated proportion zeros

```{r echo = F}
tidy(drinks_zip, type = "zero") %>% kable(digits = 3)
```

.question[
Based on the model...

- What is the probability a first-year student is a non-drinker?
- What is the probability a upperclass student (sophomore, junior, senior) is a non-drinker?

[Click here](https://forms.gle/Uh2WjEduh9Cv6KiU9) to submit your response.
]

```{r echo = F}
library(countdown)
countdown(minutes = 2, seconds = 0,
          margin = "5%")
```

---

## These are just a few of the many models...

- Use the [Vuong Test](https://bookdown.org/roback/bookdown-BeyondMLR/ch-poissonreg.html#the-vuong-test-optional) to compare the fit of the ZIP model to a regular Poisson model
  - Why can't we use a drop-in-deviance test?

- We've just discussed the ZIP model here, but there are other model applications beyond the standard Poisson regression model (e.g., hurdle models, Zero-inflated Negative Binomial models, etc. )

---

class: middle, inverse

## Likelihoods for Poisson models 

---

## Estimating coefficients in Poisson model 

- Least squares estimation would not work because the normality and equal variance assumptions don't hold for Poisson regression 

- **Maximum likelihood estimation** is used to estimate the coefficients of Poisson regression. 

- The likelihood is the product of the probabilities for the $n$ *<u>independent</u>* observations in the data 


---

## Likelihood for regular Poisson model

Let's go back to example about [household size](https://sta310-sp22.github.io/slides/06-poisson-pt1/06-poisson-pt1.html#1) in the Philippines. We will focus on the model using the main effect of age to understand variability in mean household size. 

--

Suppose the first five observations have household sizes of 4, 2, 8, 6, and 1. Then the likelihood is 

$$L = \frac{e^{-\lambda_1}\lambda_1^4}{4!} * \frac{e^{-\lambda_2}\lambda_2^2}{2!} * \frac{e^{-\lambda_3}\lambda_3^8}{8!} * 
\frac{e^{-\lambda_4}\lambda_4^6}{6!} * \frac{e^{-\lambda_5}\lambda_5^1}{1!}$$

---

## Likelihood for regular Poisson model


We will use the log likelihood to make finding the MLE easier 

$$\begin{aligned}\log(L) &= -\lambda_1 + 4\log(\lambda_1) - \lambda_2 + 2\log(\lambda_2) - \lambda_3 + 8\log(\lambda_3)\\ & -\lambda_4 + 6 \log(\lambda_4) - \lambda_5 + \log(\lambda_5) + C \end{aligned}$$

where 
  - $\lambda$ is the mean number in household depending on $x_i$
  - $C = -[\log(4!) + \log(2!) + \log(8!) + \log(6!)+ \log(1!)]$

---

## Likelihood for regular Poisson model

Given the age of the head of the household, we fit the model 

$$\log(\lambda_i) = \beta_0 + \beta_1~age_i$$

Then we replace each $\lambda_i$ in $\log(L)$ with $e^{\beta_0 + \beta_1~age_i}$. 

--

Suppose the first five observations have ages $X = (32, 21, 55, 44, 28)$. Then

--

$$\begin{aligned} \log(L) &= [-e^{\beta_0 + \beta_132}+ 4(\beta_0 + \beta_1 32)] + [ - e^{\beta_0 + \beta_121} + 2(\beta_0 + \beta_121)] \\ &+  [- e^{\beta_0 + \beta_155} + 8(\beta_0 + \beta_155)] +  [-e^{\beta_0 + \beta_144} + 6(\beta_0 + \beta_144)] \\ &+ [-e^{\beta_0 + \beta_128}(\beta_0 + \beta_128)] + C \end{aligned}$$

--

Use search algorithm to find the values of $\beta_0$ and $\beta_1$ that maximize the above equation. 

---

## Probabilities under ZIP model

There are three different types of observations in the data:

- Observed zero and will always be 0 (true zeros)
- Observed 0 but will not always be 0
- Observed non-zero count and will not always be 0

---

## Probabilities under ZIP model

**True zeros**
$$P(0 | \text{true zero})= \alpha$$

--

**Observed 0 but will not always be 0**

$$P(0 | \text{not always zero}) = (1 - \alpha)\frac{e^{-\lambda}\lambda^0}{0!}$$

--

**Did not observe 0 and will not always be 0**

$$P(z_i | \text{not always zero}) = (1 - \alpha)\frac{e^{-\lambda}\lambda^{z_i}}{z_i!}$$

---

## Probabilities under ZIP model

Putting this all together. Let $y_i$ be an observed response then 

$$P(Y_i = y_i | x_i) = \begin{cases}
\alpha + (1 - \alpha)e^{-\lambda_i} && \text{ if } y_i = 0 \\
(1 - \alpha)\frac{e^{-\lambda_i}\lambda_i^{y_i}}{y_i!} && \text{ if } y_i > 0
\end{cases}$$

--

Recall from our example, 

$$\lambda_i = e^{\beta_0 + \beta_1~off\_campus_i + \beta_2 ~ sex_i}$$

$$\alpha_i = \frac{e^{\beta_{0\alpha} + \beta_{1\alpha} ~ first\_year_i}}{1 + e^{\beta_{0\alpha} + \beta_{1\alpha} ~ first\_year_i}}$$

- Plug in $\lambda_i$ and $\alpha_i$ into the above equation obtain the likelihood function
---


## Acknowledgements

These slides are based on content in [BMLR](https://bookdown.org/roback/bookdown-BeyondMLR/ch-beyondmost.html)
 - Section 4.4.5: Using likelihoods to fit models 
 - Section 4.4.10 Case Study: Weekend Drinking



