---
title: "Finding the MLEs for Model 2"
output: pdf_document
---

This document covers multiple ways to find the MLE for Model 2 the conditional model from [Lecture 04: Using likelihoods](../04-likelihoods/04-likelihoods.html) and [Lecture 05: Using likelihoods to compare models](05-compare-models.html). See Lecture 04 for more details about the set up of the model. 

```{r warning = F, message = F}
library(tidyverse)
refs <- read_csv("data/04-refs.csv")
```

The likelihood is 

$$\begin{aligned}Lik(p_{H| N}, p_{H|H Bias}, p_{H |V Bias}) &= [(p_{H| N})^{25}(1 - p_{H|N})^{23}(p_{H| H Bias})^8 \\ &(1 - p_{H| H Bias})^{12}(p_{H| V Bias})^{13}(1-p_{H|V Bias})^9]\end{aligned}$$  
The log-likelihood is 

$$\begin{aligned}\log (Lik(p_{H| N}, p_{H|H Bias}, p_{H |V Bias})) &= 25 \log(p_{H| N}) + 23 \log(1 - p_{H|N}) \\ & + 8 \log(p_{H| H Bias}) + 12 \log(1 - p_{H| H Bias})\\ &+ 13 \log(p_{H| V Bias}) + 9 \log(1-p_{H|V Bias})\end{aligned}$$   


We would like to find the MLEs for $p_{H| N}, p_{H|H Bias}, \text{ and }p_{H |V Bias}$. 

## Finding MLEs using graphs 

We need to find the MLEs for three parameters, therefore we would need to visualize a 4-dimensional object to find the MLEs from a graph. Given the difficulty of this taks and the lack of precision in the estimates from this method, we should rely on other approaches to find the MLEs.

## Finding MLEs using calculus

We can find the MLE for each parameter using the partial derivative of the log-likelihood with respect to each parameter. 

To find the MLE for $p_{H| N}$:

$$\begin{aligned}\frac{\partial \log(Lik(p_{H| N}, p_{H|H Bias}, p_{H |V Bias}))}{\partial p_{H|N}} &= \frac{25}{p_{H|N}} - \frac{23}{1 - p_{H|N}} = 0\\[5pt] 
&\Rightarrow \frac{25}{p_{H|N}} = \frac{23}{1 - p_{H|N}} \\[5pt] 
&\Rightarrow 23p_{H|N} = 25(1 - p_{H|N}) \\[5pt] 
&\Rightarrow 48p_{H|N} = 25 \\[5pt] 
&\Rightarrow \hat{p}_{H|N} = \frac{25}{48} = 0.521\end{aligned}$$

We can use a similar approach to find the MLEs for $p_{H|H Bias}$ and $p_{H|V Bias}$. 

$$\hat{p}_{H|H Bias} = \frac{8}{20} = 0.4$$
$$\hat{p}_{H|V Bias} = \frac{13}{22} = 0.591$$

## Finding the MLEs using R

We can write a function and do a grid search to find the values that maximize the log-likelihood.

```{r}
maxloglik<- function(nvals){
  #nvals specifies the number of values
  phn <- seq(0, 1, length = nvals)
  phh <- seq(0, 1, length = nvals)
  phv <- seq(0, 1, length = nvals)
  
  loglik <- expand.grid(phn, phh, phv) 
  colnames(loglik) <- c("phn", "phh", "phv")
  
  loglik <- loglik %>%
    mutate(loglik  = log(phn^25 * (1 - phn)^23 * phh^8 * (1 - phh)^12 * 
                           phv^13 * (1 - phv)^9))
  
  loglik %>%
    arrange(desc(loglik)) %>%
    slice(1)
}
```

```{r}
maxloglik(100)
```

Depending on the number of parameters, it may be hard to test enough values for a granular enough search to find the exact values of the MLEs. Therefore, one could use the function above to conduct a crude search to find starting values for R's `optim` function. The function `optim` differs from `optimize` in that it can optimize over multiple parameter values. 

```{r}
# Function to calculate log-likelihood that will be used in the optim function
loglik <- function(params){
  phn <- params[1]
  phh <- params[2]
  phv <- params[3]

  log(phn^25 * (1 - phn)^23 * phh^8 * (1 - phh)^12 * 
                           phv^13 * (1 - phv)^9)
}
```


```{r}
# use manual search to get starting values 
start_vals <- maxloglik(50) %>% select(-loglik)
```

```{r}
# Use optim function in R to find the values to maximize the log-likelihood
optim(par = start_vals, fn = loglik, control=list(fnscale=-1))
```
