---
title: "Multilevel Model for Multinomial Response Variable"
date: "04.18.22"
output:
  xaringan::moon_reader:
    #mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta310-slides.css"
    logo: sta310-sticker.png
    lib_dir: libs/font-awesome
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%" 
      ratio: "16:9"
    self_contained: true
---

```{r setup, include = F}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(GGally)
library(xaringanExtra)
library(knitr)
library(patchwork)
library(viridis)
library(ggfortify)
library(kableExtra)
library(lme4)
library(broom.mixed)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```


---


## Announcements

- Final project due on GitHub **Wed, Apr 27 at 11:59pm**

- Please fill out course evaluations! 

- Office hours
  - Fri, Apr 22 is the last day of office hours

- Project Meetings (15 minute time slots)
  - Wed, Apr 20: 3 - 5pm 
  - Mon, Apr 25: 12 - 1pm, 3 - 5pm 
  - [Click here](https://docs.google.com/spreadsheets/d/1qmvJ435X8h0o07m16-4nznB1h6HwSnYsHOZ_hKIXMf4/edit?usp=sharing) to sign up. Must sign up at least one hour before the time block starts!

---

## Project 

- If you submitted a draft you will see feedback in GitHub by end of day today
   - Issue containing current rubric scoring and grade
   - written-report-comments.pdf document with more detailed feedback 
   - Reply to the issue to indicate if you want to keep the current grade or resubmit

- **Tips**: 
  - Clearly explain and justify the modeling choices based on analysis, previous research, subject matter expertise, etc. 
  - Make sure model equations and output from final model are consistent 
  - You can include additional output / work in the appendix. Point the reader to the appendix in your narrative.

---

class: middle,

## Open `lecture-27.Rmd` to walk through multilevel model for Multinomial Response Variable


---

class: middle, center

## `r emo::ji("tada")` Congratulations on completing STA 310! `r emo::ji("tada")`

