---
title: "Poisson Regression"
subtitle: "Offset & Zero-inflated Poisson models"
author: "Prof. Maria Tackett"
date: "01.31.22"
output:
  xaringan::moon_reader:
    #mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta310-slides.css"
    logo: sta310-sticker.png
    lib_dir: libs/font-awesome
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%" 
      ratio: "16:9"
    self_contained: true
---

```{r setup, include = F}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "70%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(GGally)
library(xaringanExtra)
library(knitr)
library(patchwork)
library(viridis)
library(ggfortify)
library(gridExtra)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: middle, center

##[Click for PDF of slides](08-poisson-pt3.pdf)

---

## Announcements

- Reading: [BMLR - Chapter 4 Poisson regression](https://bookdown.org/roback/bookdown-BeyondMLR/ch-beyondmost.html)
  -  For Monday: BMLR - [Chapter 5: Generalized Linear Models: A Unifying Theory](https://bookdown.org/roback/bookdown-BeyondMLR/ch-glms.html)


- [Mini-Project 01](https://sta310-sp22.github.io/assignments/projects/mini-project-01.html): 
  - Final write up and presentations **Wed, Feb 09 at 3:30pm**

- [HW 02](https://sta310-sp22.github.io/assignments/hw/hw-02.html) **due Mon, Feb 07 at 11:59pm**

---

## Learning goals 

- Explore properties of negative binomial versus Poisson response 

- Fit and interpret models with offset to adjust for differences in sampling effort

- Fit and interpret Zero-inflated Poisson models

---

class: middle, inverse

## Negative binomial regression model 

---

## Negative binomial regression model 

Another approach to handle overdispersion is to use a **negative binomial regression model** 

- This has more flexibility than the quasi-Poisson model, because there is a new parameter in addition to $\lambda$

<br> 

--

Let $Y$ be a **negative binomial random variable**, $Y\sim NegBinom(r, p)$, then 

$$P(Y = y_i) = {y_i + r - 1 \choose r - 1}(1-p)^{y_i}p^r \hspace{5mm} y_i = 0, 1, 2, \ldots, \infty$$

---

## Negative binomial regression model 

- **Main idea**: Generate a $\lambda$ for each observation (household) and generate a count using the Poisson random variable with parameter $\lambda$ 
  - Makes the counts more dispersed than with a single parameter 
  
- Think of it as a Poisson model such that $\lambda$ is also random 
--

.eq[
$$\begin{aligned} &\text{If }Y|\lambda \sim Poisson(\lambda)\\
&\text{ and } \lambda \sim Gamma\bigg(r, \frac{1-p}{p}\bigg)\\
&\text{ then } Y \sim NegBinom(r, p)\end{aligned}$$
]

---

## Negative binomial simulation exercise

.question[
Complete the Negative binomial regression exercise in `lecture-07.Rmd` (found in your lecture-07 GitHub repo).
]

```{r echo = F}
library(countdown)
countdown(minutes = 8, seconds = 0,
          margin = "5%")
```

---

class: middle, inverse

## Offset

---

## Data: Airbnbs in NYC

The data set [NYCairbnb-sample.csv](data/NYCairbnb-sample.csv) contains information about a random sample of 1000 Airbnbs in New York City. It is a subset of the data on 40628 Airbnbs scraped by Awad et al. (2017).

**Variables**

- `number_of_reviews`: Number of reviews for the unit on Airbnb (proxy for number of rentals)
- `price`: price per night in US dollars
- `room_type`: Entire home/apartment, private room, or shared room
- `days`: Number of days the unit has been listed (date when info scraped - date when unit first listed on Airbnb)


.footnote[Data set pulled from BMLR Section 4.11.3.]

---

## Data: Airbnbs in NYC

```{r}
airbnb <- read_csv("data/NYCairbnb-sample.csv") %>%
  select(id, number_of_reviews, days, room_type, price)
```

```{r echo = F}
set.seed(020322)
airbnb %>% sample_n(5) %>%
  kable()
```

**Goal**: Use the price and room type of Airbnbs to describe variation in the number of reviews (a proxy for number of rentals).

---

## EDA

.pull-left[
```{r echo = F, out.width = "150%"}
p1 <- ggplot(data = airbnb, aes(x = number_of_reviews)) + 
  geom_histogram() + 
   labs(x = "Number of reviews",
    title = "Distribution of number of reviews")

p2 <- airbnb %>%
  filter(price <= 2000) %>%
  group_by(price) %>%
  summarise(log_mean = log(mean(number_of_reviews))) %>%
  ggplot(aes(x = price, y = log_mean)) + 
  geom_point(alpha= 0.7) + 
  geom_smooth() + 
  labs(x = "Price in  US dollars",
    y = "Log mean number of reviews", 
    title = "Log mean number of reviews vs. price", 
    subtitle = "Airbnbs $2000 or less")

p3 <- airbnb %>%
  filter(price <= 500) %>%
  group_by(price) %>%
  summarise(log_mean = log(mean(number_of_reviews))) %>%
  ggplot(aes(x = price, y = log_mean)) + 
  geom_point(alpha= 0.7) + 
  geom_smooth() + 
  labs(x = "Price in  US dollars",
    y = "Log mean number of reviews", 
    title = "Log mean number of reviews vs. price", 
    subtitle = "Airbnbs $500 or less")

p1  / (p2 + p3) + plot_layout(heights = c(20,20))

```
]

.pull-right[

**Overall** 

```{r echo = F}
airbnb %>%
  summarise(mean = mean(number_of_reviews), 
            var = var(number_of_reviews)) %>%
  kable(digits = 3)
```

**by Room type**

```{r, echo = F}
airbnb %>%
  group_by(room_type) %>%
  summarise(mean = mean(number_of_reviews), 
            var = var(number_of_reviews)) %>%
  kable(digits = 3)
```
]

---

## Considerations for modeling

We would like to fit the Poisson regression model

$$\log(\lambda) = \beta_0 + \beta_1 ~ price + \beta_2 ~ room\_type1 + \beta_3 ~ room\_type2$$

--

.question[
- Based on the EDA, what are some potential issues we may want to address in the model building? 

- Suppose any model fit issues are addressed. What are some potential limitations to the conclusions and interpretations from the model? 
]

```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0,
          margin = "5%")
```

---

## Offset 

- Sometimes counts are not directly comparable because the observations differ based on some characteristic directly related to the counts, i.e. the *sampling effort*.

- An **offset** can be used to adjust for differences in sampling effort. 

--

- Let $x_{offset}$ be the variable that accounts for differences in sampling effort, then $\log(x_{offset})$ will be added to the model.

$$\log(\lambda_i) = \beta_0 + \beta_1 ~ x_{i1} + \beta_2 ~ x_{i2} + ... + \beta_p ~ x_{ip} + \log(x_{offset_i})$$

- The offset is a term in the model with coefficient always equal to 1.

---

## Adding an offset to the Airbnb model 

We will add the offset $\log(days)$ to the model. This accounts for the fact that we would expect Airbnbs that have been listed longer to have more reviews. 

$$\log(\lambda_i) = \beta_0 + \beta_1 ~ price_i + \beta_2 ~ room\_type1_i + \beta_3 ~ room\_type2_i + \log(days_i)$$
<br>


**Note:** The response variable for the model is still  $\log(\lambda_i)$, the log mean number of reviews

---

## Detail on the offset

We want to adjust for the number of days, so we are interested in $\frac{reviews}{days}$.

--

Given $\lambda$ is the mean number of reviews

$$\log\Big(\frac{\lambda}{days}\Big) = \beta_0 + \beta_1 ~ price + \beta_2 ~ room\_type1 + \beta_3 ~ room\_type2$$

--

$$\Rightarrow \log({\lambda}) - \log(days) = \beta_0 + \beta_1 ~ price + \beta_2 ~ room\_type1 + \beta_3 ~ room\_type2$$

--

$$\Rightarrow \log({\lambda}) = \beta_0 + \beta_1 ~ price + \beta_2 ~ room\_type1 + \beta_3 ~ room\_type2 + \log(days)$$

---

## Airbnb model in R

```{r}
airbnb_model <- glm(number_of_reviews ~ price + room_type, 
                    data = airbnb, family = poisson, 
                    offset = log(days)) #<<
```

```{r echo = F}
tidy(airbnb_model) %>%
  kable(digits = 4)
```

--

The coefficient for $\log(days)$ is fixed at 1, so it is not in the model output. 


---

## Interpretations

```{r echo = F}
tidy(airbnb_model) %>%
  kable(digits = 4)
```

<br> 

.question[
- Interpret the coefficient of `price`. 
- Interpret the coefficient of  `room_typePrivate room`
]

```{r echo = F}
library(countdown)
countdown(minutes = 3, seconds = 0,
          margin = "5%")
```


---

## Goodness-of-fit

$$\begin{aligned}&H_0: \text{ The model is a good fit for the data}\\
&H_a: \text{ There is significant lack of fit}\end{aligned}$$

```{r}
pchisq(airbnb_model$deviance, airbnb_model$df.residual, lower.tail = F)
```

--

There is evidence of significant lack of fit in the model. Therefore, more models would need to be explored that address the issues mentioned earlier. 

--

*In practice we would assess goodness-of-fit and finalize the model <b>before</b> any interpretations and conclusions.*

---

class: middle, inverse

## Zero-inflated Poisson model

---

## Data: Weekend drinking 

The data [`weekend-drinks.csv`](data/weekend-drinks.csv) contains information from a survey of 77 students in a introductory statistics course on a dry campus. 

**Variables** 

- `drinks`: Number of drinks they had in the past weekend 
- `off_campus`: 1 - lives off campus, 0 otherwise
- `first_year`: 1 - student is a first-year, 0 otherwise
- `sex`: f - student identifies as female, m - student identifies as male 

**Goal**: The goal is explore factors related to drinking behavior on a dry campus.

.footnote[Case study in BMLR - Section 4.10]

---

## EDA: Response variable

```{r, echo = F}
drinks <- read_csv("data/weekend-drinks.csv")
```


```{r echo = F, out.width = "70%"}
ggplot(data = drinks, aes(x = drinks)) + 
  geom_histogram(aes(x = drinks), binwidth = 1, alpha = 0.8, fill = "gray", color = "gray") + 
  labs(title = "Observed number of drinks", 
       subtitle = "Mean = 2.013")
```

```{r echo = F, eval = F}
drinks %>%
  summarise(mean = mean(drinks), var = var(drinks)) %>%
  kable(digits = 3)
```

---

## Observed vs. expected response

```{r echo = F,  out.width = "70%"}
drinks %>%
  mutate(sim = rpois(77, 2.013)) %>%
ggplot() +   
  geom_histogram(aes(x = drinks), binwidth = 1, alpha = 0.7, fill = "gray", color = "gray") + 
  geom_histogram(aes(x = sim), binwidth = 1, fill = "steelblue", alpha = 0.73, color = "steelblue") + 
  labs(title = "Observed (gray) vs. Expected (blue) in Poisson(2.013)")
```

--

.center[**What does it mean to be a "zero" in this data?**]

---

## Two types of zeros

There are two types of zeros

- Those who happen to have a zero in the data set (people who drink but happened to not drink last weekend)
- Those who will always report a value of zero (non-drinkers)
   - These are called **true zeros** 

--

We introduce a new parameter $\alpha$ for the proportion of true zeros,  then fit a model that has two components:

--


`r emo::ji("one")` The association between mean number of drinks and various characteristics among those who drink 

`r emo::ji("two")` The estimated proportion of non-drinkers

---

## Acknowledgements

These slides are based on content in [BMLR - Chapter 4 Poisson regression ](https://bookdown.org/roback/bookdown-BeyondMLR/ch-beyondmost.html)

