---
title: "Using Likelihoods"
author: "Prof. Maria Tackett"
date: "01.19.22"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta310-slides.css"
    logo: sta310-sticker.png
    lib_dir: libs/font-awesome
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
      slideNumberFormat: "%current%" 
      ratio: "16:9"
---

```{r setup, include = F}
knitr::opts_chunk$set(warning = FALSE, 
                      message = FALSE, 
                      fig.width = 8,
                      fig.asp = 0.618, 
                      fig.retina = 3, 
                      dpt = 300, 
                      out.width = "90%",
                      fig.align = "center")

ggplot2::theme_set(ggplot2::theme_bw(base_size = 16))

colors <- tibble::tibble(green = "#B5BA72")
```


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(GGally)
library(xaringanExtra)
library(knitr)
library(patchwork)
library(viridis)
library(ggfortify)
```

```{r xaringan-panelset, echo=FALSE}
xaringanExtra::use_panelset()
```

class: middle, center

##[Click for PDF of slides](04-likelihoods.pdf)

---

## Announcements

- [Homework 01](https://sta310-sp22.github.io/assignments/hw/hw-01.html) due Wednesday at 11:59pm 


- Week 03 reading: [BMLR: Chapter 3 - Distribution Theory](https://bookdown.org/roback/bookdown-BeyondMLR/ch-distthry.html)

- See [syllabus](https://sta310-sp22.netlify.app/syllabus/#teaching-team--office-hours) for office hours schedule 
  - Office hours online this week
  
- Team lab tomorrow - introducing mini-project 01

---

## In-person learning `r emo::ji("mask")`

Attendance to lectures and labs is expected as long as you're healthy and not in quarantine 

**Lectures**

- If you're unable to attend class, you can watch the recording of the lecture on Panopto  (link in Sakai) 
- Ask questions on GitHub Discussions or in office hours
  

**Labs**

- Labs are not recorded
- On weeks with teamwork: If you are unable to attend lab but are able enough to participate remotely, set up a Zoom call with your teammates

---

## Class Q&A Forum: GitHub Discussions

- Class Q&A forum on [GitHub Discussions](https://github.com/sta310-sp22/discussion/discussions)
  - Place for questions about course content, assignments, etc. 
  - Only use email for personal questions (e.g., grades, illness, etc.)
  - Let Prof. Tackett know if you do not have access to forum

**Demo**

---

## Homework 01

- Notes on [variable transformations]()

- Exercise 5 (and other open-ended exercises)

<font color = "red"> Add what we're looking for in this exercise</font>

---

class: middle, inverse

## Using Likelihoods

---

## Learning goals 

- Describe the concept of a likelihood

- Construct the likelihood for a simple model

- Define the Maximum Likelihood Estimate (MLE) and use it to answer an analysis question

- Identify three ways to calculate or approximate the ML and apply these methods to find the MLE for a simple model

- Use likelihoods to compare models

---

## What is the likelihood? 

A **likelihood** is a function that tells us how likely we are to observe our data for a given parameter value (or values). 

- Unlike Ordinary Least Squares (OLS), they do not require the responses be independent, identically distributed, and normal (iidN)

- It is not the same as a probability function 

--

  - **Probability function:** Fixed parameter value(s) + input possible outcomes $\Rightarrow$ probability of seeing the different outcomes given the parameter value(s)
  - **Likelihood:** Fixed data + input possible parameter values $\Rightarrow$ probability of seeing the fixed data for each parameter value

---

## Fouls in college basketball games

The data includes 30 randomly selected NCAA men's basketball games played in the 2009 - 2010 season.

We are interested in the variables `foul1`, `foul2`, and `foul3`, which indicate which team had a foul called them for the 1st, 2nd, and 3rd fouls, respectively. 
  - `H`: Foul was called on the home team 
  - `V`: Foul was called on the visiting team
  
--

We are focusing on the first three fouls for this analysis, but this could easily be extended to include all fouls in a game. 

.footnote[.small[The dataset was derived from `basektball0910.csv` used in [BMLR Section 11.2](https://bookdown.org/roback/bookdown-BeyondMLR/ch-GLMM.html#cs:refs)]]

---

## Fouls in college basketball games

```{r}
refs <- read_csv("data/04-refs.csv")
refs %>% slice(1:5) %>% kable()
```

<br> 

We will treat the games as independent in this analysis.

---

## Different likelihood models 

**Model 1 (Unconditional Model)**: What is the probability the referees call a foul on the home team, assuming foul calls within a game are independent? 

--

**Model 2 (Conditional Conditional Model)**: 
  - Is there a tendency for the referees to call more fouls on the visiting team or home team? 
  - Is there a tendency for referees to call a foul on the team that already has more fouls? 
 
--

Which model is better?

---

### Exploratory data analysis

.pull-left[
.midi[
```{r}
refs %>%
count(foul1, foul2, foul3) %>% kable()
```
]
]

.pull-right[
There are 
- 46 total fouls on the home team
- 44 total fouls on the visiting team 
]

---

class: middle, inverse

## Model 1: Unconditional model 

### What is the probability the referees call a foul on the home team, assuming foul calls within a game are independent? 

---
## Likelihood

Let $p_H$ be the probability the referees call a foul on the home team. 

**The likelihood for a single observation** 

$$Lik(p_H) = p_H^{y_i}(1 - p_H)^{n_i - y_i}$$

(In this example, we know $n_i = 3$ for all observations.)

--

**Example**

For a single game where the first three fouls are $H, H, V$, then

$$Lik(p_H) = p_H^{2}(1 - p_H)^{3 - 2} = p_H^{2}(1 - p_H)$$

---

## Model 1: Likelihood contribution 

.midi[
| Foul1 | Foul2 | Foul3 | n | Likelihood Contribution |
|-------|-------|-------|---|:-----------------------:|
| H     | H     | H     | 3 |       $p_H^3$               |
| H     | H     | V     | 2 |        $p_H^2(1 - p_H)$                 |
| H     | V     | H     | 3 |        $p_H^2(1 - p_H)$                 |
| H     | V     | V     | 7 |       A                |
| V     | H     | H     | 7 |        B                 |
| V     | H     | V     | 1 |          $p_H(1 - p_H)^2$               |
| V     | V     | H     | 5 |         $p_H(1 - p_H)^2$                |
| V     | V     | V     | 2 |         $(1 - p_H)^3$                |
]

Fill in **A** and **B**.

---

## Model 1: Likelihood function

Because the observations (the games) are independent, the **likelihood** is

$$Lik(p_H) = \prod_{i=1}^{n}p_H^{y_i}(1 - p_H)^{3 - y_i}$$
--

We will use this function to find the **maximum likelihood estimate (MLE)**. The MLE is the value between 0 and 1 where we are most likely to see the observed data.

---

## Visualizing the likelihood

.panelset.sideways[
.panel[.panel-name[Plot]
```{r echo = F}
p <- seq(0,1, length.out = 100) #sequence of 100 values between 0 and 100
lik <- p^44 *(1 -p)^46

x <- tibble(p = p, lik = lik)
ggplot(data = x, aes(x = p, y = lik)) + 
  geom_point() + 
  geom_line() +
  labs(y = "Likelihood",
       title = "Likelihood of p_H")
```
]

.panel[.panel-name[Code]
```{r eval = F}
p <- seq(0,1, length.out = 100) #sequence of 100 values between 0 and 100
lik <- p^44 *(1 -p)^46

x <- tibble(p = p, lik = lik)
ggplot(data = x, aes(x = p, y = lik)) + 
  geom_point() + 
  geom_line() +
  labs(y = "Likelihood",
       title = "Likelihood of p_H")
```
]
]

---

class: middle

.question[
What is your best guess for the MLE, $\hat{p}_H$?

A. 0.489

B. 0.500

C. 0.511

D. 0.556
]

---

## Finding the maximum likelihood estimate 

There are three primary ways to find the MLE 

`r emo::ji("white_check_mark")` Approximate using a graph

`r emo::ji("white_check_mark")` Numerical approximation 

`r emo::ji("white_check_mark")` Using calculus

---

## Approximate MLE from a graph 

```{r echo = F, out.width = "80%"}
p <- seq(0,1, length.out = 100) #sequence of 100 values between 0 and 100
lik <- p^46 *(1 -p)^44

x1 <- tibble(p = p, lik = lik)

lik_plot <- ggplot(data = x1, aes(x = p, y = lik)) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 46/90, color = "red") + 
    labs(y = "Likelihood", 
       title = "Likelihood of p_H")

lik_plot
```

---

## Find the MLE using numerical approximation

Specify a finite set of possible values the for $p_H$ and calculate the likelihood for each value

--

```{r}
# write an R function for the likelihood
ref_lik <- function(ph) {
  ph^46 *(1 - ph)^44
}
```

--

```{r}
# use the optimize function to find the MLE
optimize(ref_lik, interval = c(0,1), maximum = TRUE)
```

---

## Find MLE using calculus

- Find the MLE by taking the first derivative of the likelihood function. 

- This can be tricky because of the Product Rule, so we can maximize the **log(Likelihood)** instead. The same value maximizes the likelihood and log(Likelihood)

--

```{r echo = F, out.width = "60%"}
loglik_plot <- ggplot(data = x1, aes(x = p, y = log(lik))) + 
  geom_point() + 
  geom_line() + 
  geom_vline(xintercept = 46/90, color = "red") + 
    labs(y = "log(Likelihood)", 
       title = "log(Likelihood) of p_H")

lik_plot + loglik_plot
```

---

## Find MLE using calculus

$$Lik(p_H) = \prod_{i=1}^{n}p_H^{y_i}(1 - p_H)^{3 - y_i}$$

--

$$\begin{aligned}\log(Lik(p_H)) &= \sum_{i=1}^{n}y_i\log(p_H) + (3 - y_i)\log(1 - p_H)\\[10pt] &= 46\log(p_H) + 44\log(1 - p_H)\end{aligned}$$
---

## Find MLE using calculus

$$\frac{d}{d p_H} \log(Lik(p_H)) = \frac{46}{p_H} - \frac{44}{1-p_H} = 0$$

--

$$\Rightarrow \frac{46}{p_H} = \frac{44}{1-p_H}$$

--

$$\Rightarrow 46(1-p) = 44p$$

--

$$\Rightarrow 46 = 90p_H$$

--

$$\hat{p}_H = \frac{46}{90} = 0.511$$

---

class: middle, inverse

## Model 2: Conditional model 

### Is there a tendency for the referees to call more fouls on the visiting team or home team? 

### Is there a tendency for referees to call a foul on the team that already has more fouls?

---

## Model 2: Likelihood contributions

- Fouls are <u>not</u> independent within each game. We will specify this dependence using conditional probabilities. 
  - Recall: $P(A|B) =$ Probability of $A$ given $B$ has occurred

--

Define new parameters:

- $p_{H|N}$: Probability refs call foul on home team given there are equal numbers of fouls on the home and visiting teams 
- $p_{H|H Bias}: Probability refs call foul on home team given there are more prior fouls on the home team
- $p_{H|V Bias}: Probability refs call foul on home team given there are more prior fouls on the visiting team 

---

## Model 2: Likelihood contributions 

.midi[
| Foul1 | Foul2 | Foul3 | n | Likelihood Contribution |
|-------|-------|-------|---|:-----------------------:|
| H     | H     | H     | 3 | $(p_{H\vert N})(p_{H\vert H Bias})(p_{H\vert H Bias}) = (p_{H\vert N})(p_{H\vert H Bias})^2$ |
| H     | H     | V     | 2 |        $(p_{H\vert N})(p_{H\vert H Bias})(1 - p_{H\vert H Bias})$                 |
| H     | V     | H     | 3 |        $(p_{H\vert N})(1 - p_{H\vert H Bias})(p_{H\vert N}) = (p_{H\vert N})^2(1 - p_{H\vert H Bias})$                 |
| H     | V     | V     | 7 |       A                |
| V     | H     | H     | 7 |        B                 |
| V     | H     | V     | 1 |          $(1 - p_{H\vert N})(p_{H\vert V Bias})(1 - p_{H\vert N}) = (1 - p_{H\vert N})^2(p_{H\vert V Bias})$               |
| V     | V     | H     | 5 |         $(1 - p_{H\vert N})(1-p_{H\vert V Bias})(p_{H\vert V Bias})$                |
| V     | V     | V     | 2 |         $\begin{aligned}&(1 - p_{H\vert N})(1-p_{H\vert V Bias})(1-p_{H\vert V Bias})\\ &=(1 - p_{H\vert N})(1-p_{H\vert V Bias})^2\end{aligned}$                |
]

Fill in **A** and **B**

---

## Likelihood function 

$$\begin{aligned}Lik(p_{H| N}, p_{H|H Bias}, p_{H |V Bias}) &= [(p_{H| N})^{25}(1 - p_{H|N})^{23}(p_{H| H Bias})^8 \\ &(1 - p_{H| H Bias})^{12}(p_{H| V Bias})^{13}(1-p_{H|V Bias})^9]\end{aligned}$$      

**(Note: The exponents sum to 90, the total number of fouls in the data)**

<br> 

--

$$\begin{aligned}\log (Lik(p_{H| N}, p_{H|H Bias}, p_{H |V Bias})) &= 25 \log(p_{H| N}) + 23 \log(1 - p_{H|N}) \\ & + 8 \log(p_{H| H Bias}) + 12 \log(1 - p_{H| H Bias})\\ &+ 13 \log(p_{H| V Bias}) + 9 \log(1-p_{H|V Bias})\end{aligned}$$   

---

class: middle 

.question[
If fouls within a game are independent, how would you expect $\hat{p}_H$, $\hat{p}_{H\vert H Bias}$ and $\hat{p}_{H\vert V Bias}$ to compare? 

a. $\hat{p}_H$ is greater than $\hat{p}_{H\vert H Bias}$ and $\hat{p}_{H\vert V Bias}$

b. $\hat{p}_{H\vert H Bias}$ is greater than $\hat{p}_H $ and $\hat{p}_{H\vert V Bias}$

c. $\hat{p}_{H\vert V Bias}$ is greater than $\hat{p}_H $ and $\hat{p}_{H\vert H Bias}$

d. They are all approximately equal.

]

---

class: middle 

.question[
If there is a tendency for referees to call a foul on the team that already has more fouls, how would you expect $\hat{p}_H$ and  $\hat{p}_{H\vert H Bias}$ to compare? 

a. $\hat{p}_H$ is greater than  $\hat{p}_{H\vert H Bias}$ 

b. $\hat{p}_{H\vert H Bias}$ is greater than $\hat{p}_H$

c. They are approximately equal.
]

---


## Acknowledgements

These slides are based on content in [BMLR Chapter 2 - Beyond Least Squares: Using Likelihoods](https://bookdown.org/roback/bookdown-BeyondMLR/ch-beyondmost.html)




<!--

class: middle, inverse

## Using likelihoods to compare models 

---

## Model comparisons 

- Nested models 

- Nonnested models 

---

class: middle, inverse

## Properties of likelihoods

---

## Statistical properties (maybe take this out? )


---

## Looking ahead


- We need to add covariates to this! 

---

-->



